# Support vector machines {#svm}

## Introduction



<!-- Matt -->

<!--REPEAT CLASSIFICATION EXAMPLE FOR SERUM PROTEOMICS -->

<!-- regression and classification -->

## Classification
Load required libraries
```{r echo=T}
library(caret)

```


Setup parallel processing
```{r echo=T}
library(doMC)
registerDoMC()
getDoParWorkers()
```

MUST GENERATE A LIST OF SEEDS IF WE USE PARALLEL PROCESSING, FOR REPRODUCIBILITY

We will return to the cell segmentation data set [@Hill2007] we looked at earlier.
```{r echo=T}
data(segmentationData)
segClass <- segmentationData$Class
segData <- segmentationData[,4:61]
set.seed(42)
trainIndex <- createDataPartition(y=segClass, times=1, p=0.5, list=F)
segDataTrain <- segData[trainIndex,]
segDataTest <- segData[-trainIndex,]
segClassTrain <- segClass[trainIndex]
segClassTest <- segClass[-trainIndex]

transformations <- preProcess(segDataTrain, 
                              method=c("YeoJohnson", "center", "scale", "corr"),
                              cutoff=0.75)
segDataTrain <- predict(transformations, segDataTrain)
```


We will pass the twoClassSummary function into model training through **trainControl**. Additionally we would like the model to predict class probabilities so that we can calculate the ROC curve, so we use the **classProbs** option.
```{r echo=T}
cvCtrl <- trainControl(method = "repeatedcv", repeats = 3,
                        summaryFunction = twoClassSummary,
                        classProbs = TRUE)
```

Tune SVM over the cost parameter
```{r echo=T}
svmTune <- train(x = segDataTrain,
                 y = segClassTrain,
                 method = "svmRadial",
                 # The default grid of cost parameters go from 2^-2,
                 # 0.5 to 1,
                 # Well fit 9 values in that sequence via the tuneLength
                 # argument.
                 tuneLength = 9,
                 ## Also add options from preProcess here too
                 preProc = c("center", "scale"),
                 metric = "ROC",
                 trControl = cvCtrl)

svmTune

```


```{r echo=T}
svmTune$finalModel
```


SVM accuracy profile

```{r svmAccuracyProfile, fig.cap='SVM accuracy profile.', out.width='80%', fig.asp=0.7, fig.align='center', echo=T}
plot(svmTune, metric = "ROC", scales = list(x = list(log =
2)))
```

Test set results

```{r echo=T}
segDataTest <- predict(transformations, segDataTest)
svmPred <- predict(svmTune, segDataTest)
confusionMatrix(svmPred, segClassTest)
```

## Serum proteomics
```{r echo=T}
centre1 <- read.csv("data/serum_proteomics/male_centre1.csv")
centre2 <- read.csv("data/serum_proteomics/male_centre2.csv")

diag_cent1 <- centre1$Diagnostic_group
prot_cent1 <- centre1[,2:18]

diag_cent2 <- centre2$Diagnostic_group
prot_cent2 <- centre2[,2:18]

# featurePlot(x=prot_cent1, y=diag_cent1, plot="pairs")

transformations <- preProcess(prot_cent1, 
                              method=c("center", "scale"),
                              cutoff=0.75)
prot_cent1 <- predict(transformations, prot_cent1)

svmTune <- train(x = prot_cent1,
                 y = diag_cent1,
                 method = "svmRadial",
                 # The default grid of cost parameters go from 2^-2,
                 # 0.5 to 1,
                 # Well fit 9 values in that sequence via the tuneLength
                 # argument.
                 tuneLength = 9,
                 ## Also add options from preProcess here too
                 preProc = c("center", "scale"),
                 metric = "ROC",
                 trControl = cvCtrl)

svmTune

svmTune$finalModel

```

<!--GENERATE TWO  DIFFERENT MODELS AND COMPARE -->
LINEAR
RBM


<!--REPEAT CLASSIFICATION EXAMPLE FOR SERUM PROTEOMICS -->

## Exercises

Solutions to exercises can be found in appendix \@ref(solutions-svm)
