# Solutions ch. 4 - Linear and non-linear (logistic) regression {#solutions-logistic-regression}

Solutions to exercises of chapter \@ref(logistic-regression).

Excercise 3.1.



```{r echo=T}
genenames <- colnames(D)
geneindex <- 36
D <- read.csv(file = "/Users/christopher_penfold/Desktop/MLCourse/intro-machine-learning/data/Arabidopsis/Arabidopsis_Botrytis_transpose.csv", header = TRUE, sep = ",", row.names=1)
Xs <- seq(from = 2, to = 48, by = 2)
plot(Xs,(D[1:24,geneindex]),type="p",col="black",ylim=c(min(D[,geneindex])-0.2, max(D[,geneindex]+0.2)),main=genenames[geneindex])
points(Xs,(D[25:nrow(D),geneindex]),type="p",col="red")
```

Excercise 3.2. 

```{r echo=T}
geneindex <- 10
lrfit3 <- train(y~., data=data.frame(x=D[1:24,3:10],y=D[1:24,geneindex]), method = "lm")
```

Excercise 3.3. Linear regression can generally be applied for any number of variables. A notable example, would be to regress the expression pattern of a gene against putative regulators.

Excercise 3.4 (optional): 

Example covariance functions implemented from the [Kernel Cookbook](http://www.cs.toronto.edu/~duvenaud/cookbook/). Here we implement a rational quadratic covariance function:

```{r echo=T}
covRQ <- function(X1,X2,l=1,sig=1,a=2) {
  K <- matrix(rep(0, length(X1)*length(X2)), nrow=length(X1))
  for (i in 1:nrow(K)) {
    for (j in 1:ncol(K)) {
      K[i,j] <- sig^2*(1 + (abs(X1[i]-X2[j])^2/(2*a*l^2))    )^a 
    }
  }
  return(K)
}
```

Here we implement a periodic covariance function:

```{r echo=T}
covPer <- function(X1,X2,l=1,sig=1,p=1) {
  K <- matrix(rep(0, length(X1)*length(X2)), nrow=length(X1))
  for (i in 1:nrow(K)) {
    for (j in 1:ncol(K)) {
      K[i,j] <- sig^2*exp(sin(pi*abs(X1[i]-X2[j])/p)^2 / l^2) 
    }
  }
  return(K)
}
```

Excercise 3.5:

We can example functions using the following code:

```{r echo=F}
require(MASS)
require(plyr)
require(reshape2)
require(ggplot2)

covSE <- function(X1,X2,l=1,sig=1) {
  K <- matrix(rep(0, length(X1)*length(X2)), nrow=length(X1))
  for (i in 1:nrow(K)) {
    for (j in 1:ncol(K)) {
      K[i,j] <- sig^2*exp(-0.5*(abs(X1[i]-X2[j]))^2 /l^2)
    }
  }
  return(K)
}
```

```{r echo=T}
x.star <- seq(-5,5,len=500)
f <- data.frame(x=c(-4,-3,-2,-1,0,1,2),
                y=sin(c(-4,-3,-2,-1,0,1,2)))
x <- f$x
k.xx <- covSE(x,x)
k.xxs <- covSE(x,x.star)
k.xsx <- covSE(x.star,x)
k.xsxs <- covSE(x.star,x.star)

f.star.bar <- k.xsx%*%solve(k.xx)%*%f$y  #Mean
cov.f.star <- k.xsxs - k.xsx%*%solve(k.xx)%*%k.xxs #Var

y1 <- mvrnorm(1, f.star.bar, cov.f.star)
y2 <- mvrnorm(1, f.star.bar, cov.f.star)
y3 <- mvrnorm(1, f.star.bar, cov.f.star)
plot(x.star,sin(x.star),type = 'l',col="red",ylim=c(-2.2, 2.2))
points(f,type = 'p',col="blue")
lines(x.star,y1,type = 'l',col="blue")
lines(x.star,y2,type = 'l',col="blue")
lines(x.star,y3,type = 'l',col="blue")
```


## Exercise 2


